### 2017-07-08 paper/extensions.tex
small correction in Remark 2
OK

As in the case of a.s. termination we start with general mathematical statement 
about LexRSMs. We define a restricted class of strict LexRSMs with \emph{bounded 
expected conditional increase} property. Recall from 
definition~\ref{def:lexrsm} that strict
LexRSM for a stopping time $\stime$ is characterized by the possibility to 
a.s. partition, for each $i\in \Nset_0$, the set $\{\omega\in \Omega\mid 
\stime(\omega)>i \}$ into $n$ sets $L^i_1,\dots,L^i_d$ such that, intuitively, 
on $L^i_j$ the conditional expectation of $\vec{X}_{i+1}[j]$ given $\genfilt_i$ 
is smaller than $\vec{X}_i[j]$, and for all $j'<j$, on $L^i_{j'}$ the 
conditional expectation of $\vec{X}_{i+1}[j]$ given $\genfilt_i$ 
is no larger than $\vec{X}_i[j]$. This leaves the opportunity of conditional 
expectation of $\vec{X}_{i+1}[j]$ being larger than $\vec{X}_i[j]$ on 
$L^i_{j''}$ with $j''>j$. The expected conditional increase (ECI) property 
bounds the 
possibility of this increase.

@@

Previous work testifies that analogous ``bounded 
expected change'' constraints naturally emerge when one aims to obtain 
additional information about runtime of programs 
~\cite{HolgerPOPL,CFNH16:prob-termination,CNZ17,CF17}. The major novelty of our 
bounded ECI condition is that it can be used to obtain polynomial bounds on 
expected runtime using \emph{linear} supermartingales.

>>>>

As in the case of a.s. termination we start with general mathematical statement 
about LexRSMs. We define a restricted class of strict LexRSMs with \emph{bounded 
expected conditional increase} property. Recall from 
Definition~\ref{def:lexrsm} that strict
LexRSM for a stopping time $\stime$ is characterized by the possibility to 
a.s. partition, for each $i\in \Nset_0$, the set $\{\omega\in \Omega\mid 
\stime(\omega)>i \}$ into $n$ sets $L^i_1,\dots,L^i_d$ such that, intuitively, 
on $L^i_j$ the conditional expectation of $\vec{X}_{i+1}[j]$ given $\genfilt_i$ 
is smaller than $\vec{X}_i[j]$, and for all $j'<j$, on $L^i_{j'}$ the 
conditional expectation of $\vec{X}_{i+1}[j]$ given $\genfilt_i$ 
is no larger than $\vec{X}_i[j]$. This leaves the opportunity of conditional 
expectation of $\vec{X}_{i+1}[j]$ being larger than $\vec{X}_i[j]$ on 
$L^i_{j''}$ with $j''>j$. The expected conditional increase (ECI) property 
bounds the 
possibility of this increase.

@@

\begin{remark}
Previous works testify that analogous ``bounded expected change'' constraints 
naturally emerge when one aims to obtain additional information about termination 
time of probabilistic programs~\cite{HolgerPOPL,CFNH16:prob-termination,CNZ17}.
However all previous works consider bounded expected change to use well-known 
tail-bound inequalities for martingales such as Azuma's inequality to obtain 
probability bounds on termination time, but not asymptotic bounds on expected
termination time.
In contrast, the major novelty of our bounded ECI condition is that it we show it can be used to 
obtain asymptotic bounds on expected termination time, and in particular, they 
can be used to obtain polynomial bounds on expected runtime using \emph{linear} 
supermartingales.
\end{remark}

### 2017-07-08 paper/lexicographic-programs.tex
OK

%\textbf{[PETR: DEFINE LOCATION TERMINATION]}

@@

$[x\geq 0]$	
	\end{lstlisting}
\end{lrbox}
\begin{figure}[t]
	\centering
	\usebox{\infas}
	\hspace{0.1cm}
	\usebox{\infast}
	\hspace{0.1cm}
	\usebox{\infastinv}
\caption{An a.s. terminating program with infinite expected termination time. A 
2-dimensional $1$-LinLexRSM map for the program is given on the right, along 
with the 
supporting invariants in square brackets. The invariants and a LinLexRSM on 
each 
line belong to the program location in which the program is \emph{before} 
executing the command on that line. The function is indeed a $1$-LinLexRSM, 
since in the probabilistic branching location $\loc$ we have 
$\preexp{\lem}(\loc,(x,c))=3c+3$ and $3c+3\leq 6c+1-1$ for all $c\geq 1$.} 
\label{fig:inftime}
\end{figure}

\begin{example}
\label{ex:infinite-time}
Consider the program in Figure~\ref{fig:inftime}. It is easy to see that it 
terminates a.s., but the expected termination time is infinite: to see this, 
note that that the expected value of variable $x$ upon reaching the second loop 
is $\frac{1}{2}\cdot 1 + \frac{1}{4}\cdot 2 + \frac{1}{8}\cdot 4 + \cdots = 
\frac{1}{2}+\frac{1}{2}+\frac{1}{2}+\cdots=\infty$ and that the time needed to 
get out of the second loop is equal to the value of $x$ upon entering the loop. 
However, a.s. termination of the program is proved by a 2-dimensional LinLexRSM 
map
$\lem$ pictured in the figure.
\end{example}

>>>>

\begin{example}\label{ex:lexrsmbasic} 
Consider the program shown in Figure~\ref{fig:inftime}.
A 2-dimensional $1$-LinLexRSM map for the program is given on the right, along 
with the supporting invariants in square brackets. The invariants and a LinLexRSM 
on each line belong to the program location in which the program is \emph{before} 
executing the command on that line. The function is indeed a $1$-LinLexRSM, 
since in the probabilistic branching location $\loc$ we have 
$\preexp{\lem}(\loc,(x,c))=3c+3$ and $3c+3\leq 6c+1-1$ for all $c\geq 1$.
\end{example}
%\textbf{[PETR: DEFINE LOCATION TERMINATION]}

@@

$[x\geq 0]$	
	\end{lstlisting}
\end{lrbox}
\begin{figure}[t]
	\centering
	\usebox{\infas}
	\hspace{0.1cm}
	\usebox{\infast}
	\hspace{0.1cm}
	\usebox{\infastinv}
\caption{An example program that is a.s. terminating but with infinite expected termination time.} 
\label{fig:inftime}
\end{figure}

\begin{example}
\label{ex:infinite-time}
Consider the program in Figure~\ref{fig:inftime}. 
By Example~\ref{ex:lexrsmbasic} and Theorem~\ref{thm:lexrsm-programs} it follows that 
the program terminates a.s., but the expected termination time is infinite: to see this, 
note that that the expected value of variable $x$ upon reaching the second loop 
is $\frac{1}{2}\cdot 1 + \frac{1}{4}\cdot 2 + \frac{1}{8}\cdot 4 + \cdots = 
\frac{1}{2}+\frac{1}{2}+\frac{1}{2}+\cdots=\infty$ and that the time needed to 
get out of the second loop is equal to the value of $x$ upon entering the loop. 
%However, a.s. termination of the program is proved by a 2-dimensional LinLexRSM map
%$\lem$ pictured in the figure.
\end{example}
